---
title: "Case Study 2"
author: "Alexandra Faber, Lena Hauer, Fiona Trumshi"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
subtitle: AKSTA Statistical Computing
header-includes: \usepackage{booktabs}
---
## AI Usage

\begin{table}[ht]
\centering
\begin{tabular}{l|rrrrrrrrr}
\toprule
 \textbf{AI usage} & \textbf {Question a} & \textbf{Question b} & \textbf{Question c} &\textbf{Question d} &\textbf{Question e} & \textbf{Question f}\textbf{Question g}\textbf{Question i}\textbf{Question j}  \\
\midrule
\textit{Feedback on existing solution}  & no & no & no &  &    &  &  &  &   \\ 
\textit{Improving existing code}   & little & no & no &  &    &  &  &  &    \\ 
\textit{Generating new code}  & no & no & no &  &    &   &  &  &   \\ 
\textit{Layout and formatting}  & no & no & no &  &     &   &  &  &    \\ 
\textit{Text correction}   & no & no & no &  &  &  &  &  &        \\ 
\textit{Debugging} & no & no & no &  &  &  &  &  &   \\
\textit{AI Model} & GPT-5.3  &  &  &    &  & &  &  &   \\ 
\bottomrule
\end{tabular}
\end{table}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## a. Data Importing And Cleaning  

Load the following datasets from TUWEL and ensure that missing values are handled correctly and column
names are clear. Each dataset should ultimately contain only two columns: country and the respective
variable. Note that some data sets also contain information on the year when the value was last updated.
• rawdata_369.txt which contains the (estimated) public expenditure on education as a percent of GDP.
Pay attention! The delimiter is 2 or more white spaces (one space would linot work as it would separate
country names which contain a space); you have to skip the first two lines.
• rawdata_373.csv which contains the (estimated) youth unemployment rate (15-24) per country
• rawdata_347.txt which contains (estimated) net migration rate per country.


```{r cars}

library(dplyr)
library(tidyr)
library(magrittr)

getwd()

youthUnemployment <- read.csv('rawdata_373.csv', sep = ',', colClasses = c("character", "numeric"), col.names = c('Country', 'Youth_Unemployment_Rate'))
youthUnemployment <- youthUnemployment %>% separate(Country, into = c("Country", "Whitespace"), sep = '\\s{2,}')
youthUnemployment <- select(youthUnemployment, Country, Youth_Unemployment_Rate)


edu_expend <- read.delim('rawdata_369.txt', skip = 1, header = FALSE, sep = '^')
edu_expend <- edu_expend %>% separate(V1, into = c("Index", "Country", "Education_Expenditure", "Year" ), sep = '\\s{2,}')
edu_expend <- select(edu_expend, Country, Education_Expenditure)


net_migration <- read.delim('rawdata_347.txt', skip = 1, header = FALSE, sep = '^')

net_migration <- net_migration %>% separate(V1, into = c("Index", "Country", "Net_Migration", "Year" ), sep = '\\s{2,}')
net_migration <- select(net_migration, Country, Net_Migration)



```

## b. Merging Raw Data

Merge the datasets using dplyr on a unique key and retain the union of all observations.

What key are you using for merging?
Return the dimensions of the merged dataset.

A:

- The key country is being used, since that is the one variable that every data set contains and the variable the others are depending on. The join used to merge is the full join because it keeps all observations from all three tables.
- The dimension of the dataset is 227 4 which is 227 rows(oberservation) with 4 columns (variables). The amount of observation is equal to the amount of observations in the largest dataset (Net Migration Rate). Therefore we can conclude, that the merge was successful. 
```{r}


merged_dataset <- youthUnemployment %>% full_join(edu_expend, by = "Country") %>% full_join(net_migration, by = "Country")
dim(merged_dataset)


```

## c. Enriching Data with Income Classification

Obtain country income classification from the World Bank and merge it with the dataset.

Identify common variables between the datasets. Can they be used for merging? Why or why not?
Since ISO Codes are standardized, download and use the CIA country data codes for merging. Make sure you are not losing any of the countries in your original dataset when merging. 

```{r}
income_classification <- read.csv('Income Classification.csv', header = FALSE, sep = ";", skip = 1, col.names = c("Country", "Code", "Region", "Category", "Lending Category"))
income_classification <- select(income_classification, Category, Code)
income_classification
#merge_all <- income_classification %>% full_join(merged_dataset, by = "Country")
#merge_all
```

With CIA Country Codes:

```{r}
cia_country_codes <- read.csv('Country Data Codes.csv', col.names = c("Country", "Code", "ISO.3166", "Stanag", "Internet", "Comment"))
cia_country_codes <- select(cia_country_codes, "Country", "Code")
merge_country_coudes <- merged_dataset %>% left_join(cia_country_codes, by = "Country")
merge_country_coudes

merge_all <-merge_country_coudes %>% left_join(income_classification, by = "Code")
merge_all
```









