---
title: "Case Study 2"
author: "Alexandra Faber, Lena Hauer, Fiona Trumshi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    highlight: kate
    latex_engine: xelatex
  html_document:
    df_print: paged
subtitle: AKSTA Statistical Computing
header-includes: \usepackage{booktabs}
---
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, warning = FALSE, attr.source='.numberLines')
knitr::opts_chunk$set(fig.width=12, fig.height=10, out.width='85%', fig.align='center', warning=FALSE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```
## AI Usage

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{l|ccccccccc}
\toprule
\textbf{AI usage} & \textbf{Question a} & \textbf{Question b} & \textbf{Question c} & \textbf{Question d} & \textbf{Question e} & \textbf{Question f} & \textbf{Question g} & \textbf{Question i} & \textbf{Question j} \\
\midrule
\textit{Feedback on existing solution} & no & no & no & no & no & no & no & no & no \\
\textit{Improving existing code}       & little & no & no & little & little & little & no & little & no \\
\textit{Generating new code}           & no & no & no & no & no & no & no & no & no \\
\textit{Layout and formatting}         & no & no & no & no & no & no & no & no & no \\
\textit{Text correction}               & no & no & no & no & no & no & no & little & no \\
\textit{Debugging}                     & no & no & no & no & no & no & no & no & no \\
\textit{AI Model}                      & GPT-5.3 & & & Gemini 3.1 & Gemini 3.1 & Gemini 3.1 & no & Sonnet 4.6 & no \\
\bottomrule
\end{tabular}%
}
\end{table}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## a. Data Importing And Cleaning  

Load the following datasets from TUWEL and ensure that missing values are handled correctly and column
names are clear. Each dataset should ultimately contain only two columns: country and the respective
variable. Note that some data sets also contain information on the year when the value was last updated.\\
• rawdata_369.txt which contains the (estimated) public expenditure on education as a percent of GDP.
Pay attention! The delimiter is 2 or more white spaces (one space would linot work as it would separate
country names which contain a space); you have to skip the first two lines.\\
• rawdata_373.csv which contains the (estimated) youth unemployment rate (15-24) per country\\
• rawdata_347.txt which contains (estimated) net migration rate per country.


```{r}
library(ggplot2)
library(broom)
library(patchwork)
library(dplyr)
library(tidyr)
library(magrittr)

youthUnemployment <- read.csv('rawdata_373.csv', sep = ',', colClasses 
                              = c("character", "numeric"), 
                         col.names = c('Country', 'Youth_Unemployment_Rate'))
youthUnemployment <- youthUnemployment %>% separate(Country, into
                                = c("Country", "Whitespace"), sep = '\\s{2,}')
youthUnemployment <- select(youthUnemployment, Country, Youth_Unemployment_Rate)


edu_expend <- read.delim('rawdata_369.txt', skip = 1, header = FALSE, 
                         sep = '^')
edu_expend <- edu_expend %>% separate(V1, into = c("Index", "Country", 
                      "Education_Expenditure", "Year" ), sep = '\\s{2,}')
edu_expend <- select(edu_expend, Country, Education_Expenditure)


net_migration <- read.delim('rawdata_347.txt', skip = 1, header = FALSE, 
                            sep = '^')

net_migration <- net_migration %>% separate(V1, into = c("Index", 
                "Country", "Net_Migration", "Year" ), sep = '\\s{2,}')
net_migration <- select(net_migration, Country, Net_Migration)



```

## b. Merging Raw Data

Merge the datasets using dplyr on a unique key and retain the union of all observations.

What key are you using for merging?
Return the dimensions of the merged dataset.

Answer:

- The key "Country" is being used, since that is the one variable that every data set contains and the variable the others are depending on. The join used to merge is the full join because it keeps all observations from all three tables.
- The dimension of the dataset is 227 4 which is 227 rows(oberservation) with 4 columns (variables). The amount of observations is equal to the amount of observations in the largest dataset (Net Migration Rate). Therefore we can conclude that the merge was successful. 
```{r}


merged_dataset <- youthUnemployment %>% full_join(edu_expend, by = 
                 "Country") %>% full_join(net_migration, by = "Country")
dim(merged_dataset)


```

## c. Enriching Data with Income Classification

Obtain country income classification from the World Bank and merge it with the dataset.

Identify common variables between the datasets. Can they be used for merging? Why or why not?
Since ISO Codes are standardized, download and use the CIA country data codes for merging. Make sure you are not losing any of the countries in your original dataset when merging. 

```{r}
income_classification <- read.csv('Income Classification.csv', 
                         header = FALSE, sep = ";", skip = 1, 
                         col.names = c("Country", "Code", "Region", "Category", 
                                       "Lending Category"))
income_classification <- select(income_classification, Category, Code)
#income_classification
#merge_all <- income_classification %>% full_join(merged_dataset, by = "Country")
#merge_all
```
The income classification dataset cannot be successfully merged with our dataset since the country names are not equal, which leads to a larger dataset than wanted (some countries can be merged, others will appear twice in the new dataset). This problem can be tackled via using a dataset like the cia country codes and merging it with the others. 
With CIA Country Codes:

```{r}
cia_country_codes <- read.csv('Country Data Codes.csv', 
                     col.names = c("Country", "Code", "ISO.3166", 
                                  "Stanag", "Internet", "Comment"))
cia_country_codes <- select(cia_country_codes, "Country", "Code")
merge_country_coudes <- merged_dataset %>% left_join(cia_country_codes, 
                                                     by = "Country")

merge_all <-merge_country_coudes %>% left_join(income_classification, 
                                               by = "Code")
```



## d. Adding Geographical Information
Introduce continent and subcontinent (or region) data for each country.
Find and download an appropriate online resource.
Merge this information into the dataset, naming the final dataset df_vars. Make sure you are not losing any of the countries in your original data set when merging.

```{r }
library(dplyr)
library(readr)

# Loading the geograhical resource (Downloaded from:
#https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes/blob/master/all/all.csv)
# reading only the relevant information
geo_data <- read_csv("countries-with-regional-codes.csv", 
                     show_col_types = FALSE) %>%
  select(`alpha-3`, region, `sub-region`) %>%
  rename(code = `alpha-3`, continent = region, subcontinent = `sub-region`)

# Using left_join to ensure no countries from the original set are lost
df_vars <- merge_all %>%
  left_join(geo_data, by = c("Code" = "code"))

# Verify dimensions to ensure the number of countries stayed the same
dim(df_vars)
#df_vars
```
The dimensions are the same as before,  so no country is lost.

## e. Data Tidiness and Summary Statistics

Evaluate the tidiness of df_vars (observational units, variables, fixed vs. measured variables). Make adjustments to tidy the data, if necessary.

The dataset df_vars follows the principles of tidy data:

Each variable forms a column: Expenditure, Unemployment, and Migration are separate columns.

Each observation forms a row: Each row represents a unique country/entity.

Each type of observational unit forms a table: The unit here is the country.
Fixed variables include the Country Name and Continent, while the measured variables are the economic indicators.

Create a frequency table for the income status variable and briefly interpret the results
```{r }
# Ensure income_status is a factor with proper levels for ordering
df_vars$income_status <- factor(df_vars$Category, 
                        levels = c("Low income", "Lower middle income", 
                                   "Upper middle income", "High income"))

# Frequency table
income_table <- df_vars %>%
  filter(!is.na(income_status)) %>%
  count(income_status) %>%
  mutate(relative_freq = n / sum(n))

print(income_table)
```
Interpretation: The frequency table shows the distribution of countries across wealth tiers. The frequency distribution reveals that the majority of countries fall within the Middle/High income brackets, with the High income group being the most frequent. The relatively low percentage of 'Low income' countries (approx. 12%) suggests that while significant wealth exists globally, national-level extreme poverty is concentrated in a specific minority of entities. There are also a lot of territories in this dataset that are high-income but are not independent countries, which skews the results a bit and some data is missing (from Macedonia, for example).

Analyze the distribution of income status across continents by computing absolute and relative frequencies. Comment on the findings.

```{r}
# Absolute and relative frequencies by continent
continent_dist <- df_vars %>%
  filter(!is.na(continent), !is.na(income_status)) %>%
  group_by(continent, income_status) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(relative_freq = count / sum(count))

print(continent_dist)

# Identify unique countries (only ones in their income group in that continent)
unique_in_group <- df_vars %>%
  filter(!is.na(continent)) %>%
  group_by(continent, income_status) %>%
  filter(n() == 1) %>%
  select(Country, continent, income_status)

print(unique_in_group)
```


Using the distribution of income status across continents, identify which countries are the only ones in their income group across the continent. Discuss briefly.

Discussion: The only country completely alone in the income category in a continent are the Seychelles in Africa, which is the only High income country on that continent. This group of islands has a strong economy that largely depends on tourism and fishing.


## f. Further Summary Statistics and Insights
Create a table of average (mean and median) values for expenditure, youth unemployment rate and net migration rate separated into income status. Make sure that in the output, the ordering of the income classes is proper (i.e., L, LM, UM, H or the other way around). Briefly comment the results and any differences between the mean and median.

```{r}
# Convert columns to numeric to avoid "compatible vectors" error
df_vars <- df_vars %>%
  mutate(
    Education_Expenditure = as.numeric(Education_Expenditure),
    Youth_Unemployment_Rate = as.numeric(Youth_Unemployment_Rate),
    Net_Migration = as.numeric(Net_Migration)
  )

# Calculate Mean and Median
summary_table <- df_vars %>%
  filter(!is.na(income_status)) %>%
  group_by(income_status) %>%
  summarise(
    mean_edu   = mean(Education_Expenditure, na.rm = TRUE),
    med_edu    = median(Education_Expenditure, na.rm = TRUE),
    mean_unemp = mean(Youth_Unemployment_Rate, na.rm = TRUE),
    med_unemp  = median(Youth_Unemployment_Rate, na.rm = TRUE),
    mean_mig   = mean(Net_Migration, na.rm = TRUE),
    med_mig    = median(Net_Migration, na.rm = TRUE),
    .groups = "drop"
  )

summary_table
```
Commentary on Results:
By comparing the mean and median, we can detect skewness in the data. For instance, in the Net Migration variable, the mean is often heavily influenced by a few outlier countries with extreme migration flows, whereas the median provides a better "typical" value for that income group. The mean of the unemployment rate is also very influenced by huge outliers. For the  education expendures the median and mean values are very similiar.

Look at the standard deviation and the interquartile range of the variables per income status instead of
the location statistics above. Do you gain additional insights? Briefly comment the results.

```{r }
# 1. Education Expenditure Dispersion Table
edu_dispersion <- df_vars %>%
  filter(!is.na(income_status)) %>%
  group_by(income_status) %>%
  summarise(
    sd_education = sd(Education_Expenditure, na.rm = TRUE),
    iqr_education = IQR(Education_Expenditure, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Youth Unemployment Dispersion Table
youth_dispersion <- df_vars %>%
  filter(!is.na(income_status)) %>%
  group_by(income_status) %>%
  summarise(
    sd_unemployment = sd(Youth_Unemployment_Rate, na.rm = TRUE),
    iqr_unemployment = IQR(Youth_Unemployment_Rate, na.rm = TRUE),
    .groups = "drop"
  )

# 3. Net Migration Dispersion Table
migration_dispersion <- df_vars %>%
  filter(!is.na(income_status)) %>%
  group_by(income_status) %>%
  summarise(
    sd_migration = sd(Net_Migration, na.rm = TRUE),
    iqr_migration = IQR(Net_Migration, na.rm = TRUE),
    .groups = "drop"
  )

# Display the tables one after another
print(edu_dispersion)
print(youth_dispersion)
print(migration_dispersion)
```
Education Expenditures: 
The standard deviation and the IQR are relatively close, which means the data is more normally distributed.

Youth Unemployment rate: 
The IQR is a bit larger than the SD, which means that there are a few outliers that influence the sd greatly but not so much the IQR.

Migration: The sd is often larger than the iqr, especially for lower middle income, which has a sd that is six times its IQR. that means this group has some quite extreme outliers, which the IQR ignores but the sd doesn't.



Extend the analysis of the statistics median and IQR to each income status and continent combination. Play around with displaying the resulting table. Use pivot_longer() and/or pivot_wider()
to generate different outputs. Discuss the results as well as the readability of the different tables.


```{r}
# Long format
stats_long <- df_vars %>%
  filter(!is.na(continent), !is.na(income_status)) %>%
  group_by(continent, income_status) %>%
  summarise(across(c(Education_Expenditure, 
                     Youth_Unemployment_Rate, Net_Migration), 
                   list(med = ~median(.x, na.rm = TRUE), 
                        iqr = ~IQR(.x, na.rm = TRUE))),
            .groups = "drop") %>%
  pivot_longer(cols = -c(continent, income_status), 
               names_to = "metric", 
               values_to = "value")
stats_long
# Wide format for readability
stats_wide <- stats_long %>%
  pivot_wider(names_from = income_status, values_from = value)

stats_wide
```
The table created by pivot_wider is much more readable than the one with pivot_longer, because with the wider command the income classes are displayed beside each other, which makes comparison much easier.

Identify countries performing well in terms of both youth unemployment and net migration rate (top 25% in net migration and bottom 25% in youth unemployment within their continent).
```{r}
# Top 25% in net migration and bottom 25% in youth unemployment within continent
top_performers <- df_vars %>%
  filter(!is.na(continent)) %>%
  group_by(continent) %>%
  filter(
    Net_Migration >= quantile(Net_Migration, 0.75, na.rm = TRUE),
    Youth_Unemployment_Rate <= quantile(Youth_Unemployment_Rate, 0.25, 
                                        na.rm = TRUE)
  ) %>%
  select(Country, continent, Net_Migration, Youth_Unemployment_Rate)

top_performers
```
This data shows us that there is quite a difference between all the continents. The migration in Africa  and Oceania very small compared to the European an Asian top countries. Quatar performs especially well because it has a very low youth unemployment rate with a quite high Migration.


## g. Conditional probabilities

Estimate the following based on the observed frequencies in the data:\\
• What is the (posterior or conditional) probability that a European country belongs to the high income
group? What is the prior probability that a country belongs to the high income group?


```{r}
g_1 <- df_vars %>% filter(continent == "Europe" & !is.na(income_status)) %>% 
  count(income_status) %>% 
  mutate(perc_income_EU = round(n/sum(n) * 100, 2) ) %>% 
  filter(income_status == "High income") %>% 
  select(perc_income_EU) %>% as.numeric()

cat("Posterior or conditional probability that a European country belongs\n",
    "to the high income group ", g_1, "%", sep = "")
```

```{r}
g_2 <- df_vars %>% filter(!is.na(income_status)) %>% 
  count(income_status) %>% mutate(perc_income = round(n/sum(n)*100, 2)) %>% 
filter(income_status == "High income") %>% select(perc_income) %>% as.numeric()

cat(paste("Prior probability that a country belongs to the high income group ", g_2, "%"), sep = "")
```

• Given a country has high youth unemployment (above %25), what is the probability that it also has
negative net migration?

```{r}
g_3 <- df_vars %>% filter(Youth_Unemployment_Rate > 25 & 
                  !is.na(Youth_Unemployment_Rate) & !is.na(Net_Migration)) %>% 
  mutate(net_migration_sign = ifelse(Net_Migration < 0, "-", "+")) %>% 
  relocate(net_migration_sign, .after = Net_Migration) %>% 
  count(net_migration_sign) %>% 
  mutate(perc_net_migr_sign = round(n/sum(n)*100, 2)) %>% 
  filter(net_migration_sign == "-") %>% select(perc_net_migr_sign) %>% 
  as.numeric()
cat(paste("Given a country has high youth unemployment (above 25%),"),
    paste("the probability that it also has negative net migration is", g_3, "%"),
    sep = "\n")
```


## h. Simpson’s Paradox Analysis
Investigate whether an overall trend in youth unemployment rate in the high and low income groups reverses
when analyzed at the continent level. E.g., does the youth unemployment rate appear lower in low-income
countries overall, but higher when controlling for continent? Explain the results and possible reasons behind
this paradox.

```{r}
model_overall <- lm(Youth_Unemployment_Rate~income_status, data = df_vars)
model_continent <- lm(Youth_Unemployment_Rate~income_status+continent, 
                      data = df_vars)
summary(model_overall)
summary(model_continent)
```

Two linear models are fitted to compare the effect of high income on youth unemployment rate when continents are omitted and when they are included as control variables. The summary of the overall model shows that, compared to low income, all other income categories have a higher (positive) effect on youth unemployment rate. However, only upper middle income is significant. On the other hand, the coefficients for income status are much higher for all categories when continents are included in the model, with Africa as the reference continent. Here, high income has a significant p-value with a coefficient nearly 5 times higher than in the overall model.

```{r}

coef_overall    <- coef(model_overall)["income_statusHigh income"]
coef_controlled <- coef(model_continent)["income_statusHigh income"]

cat("Overall effect of High income: ", round(coef_overall, 3), "\n")
cat("Effect controlling for continent: ", round(coef_controlled, 3), "\n")
```

```{r}
tidy_overall    <- tidy(model_overall,    conf.int = TRUE) %>% 
  mutate(model = "Overall (no continent)")
tidy_controlled <- tidy(model_continent, conf.int = TRUE) %>% 
  mutate(model = "Controlled for continent")

bind_rows(tidy_overall, tidy_controlled) %>%
  filter(term == "income_statusHigh income") %>%
  ggplot(aes(x = model, y = estimate, color = model)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), 
                width = 0.2, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Effect of High Income on Youth Unemployment",
    x = NULL, y = "Coefficient (vs. Low income)",
    caption  = "Error bars = 95% confidence interval"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
df_plot <- df_vars %>%
  filter(!is.na(Youth_Unemployment_Rate),
         !is.na(continent), !is.na(income_status))

# Overall means
overall_means <- df_plot %>% filter(!is.na(Youth_Unemployment_Rate),
         !is.na(continent) & !is.na(income_status)) %>% 
  group_by(income_status) %>%
  summarise(mean_unemp = mean(Youth_Unemployment_Rate, na.rm = TRUE))

# Within-continent means
continent_means <- df_plot %>%
  group_by(continent, income_status) %>%
  summarise(mean_unemp = mean(Youth_Unemployment_Rate, na.rm = TRUE), 
            .groups = "drop")

p1 <- ggplot(overall_means, aes(x = income_status, y = mean_unemp, 
                                fill = income_status)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = round(mean_unemp, 1)), vjust = -0.5) +
  labs(title = "A: Overall (marginal)", y = "Mean Youth Unemployment (%)", 
       x = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p2 <- ggplot(continent_means, aes(x = income_status, y = mean_unemp, 
                                  fill = income_status)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = round(mean_unemp, 1)), 
            position = position_dodge(0.6), vjust = -0.5, size = 3) +
  facet_wrap(~ continent, nrow = 1) +
  labs(title = "B: Within each continent (controlled)", y = NULL, x = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank())

p1 + p2 + plot_layout(widths = c(1, 3)) +
  plot_annotation(
    title    = "Simpson's Paradox: Youth Unemployment by Income Group"
  )
```

The plot shows the mean youth unemployment rate overall (across all continents) and then separately for each continent. Americas, Europe, and Oceania have no countries classified as low income, therefore the mean was only computed for high income and middle income categories. While the overall trend shows that mean youth unemployment rate is higher in the upper middle income group, the trend is reversed when controlling for continent. The Americas show a higher unemployment rate for high income countries, while Asia has a large proportion of low income countries with high unemployment rates. This is due to the fact that each continent has a different distribution of income groups among its countries due to economic factors. Hence, observing a global effect alone is limiting, whereas analysing the same phenomenon across different groups can reveal reversed trends.


## i. Data Export
Export the final tidy dataset from e. as a CSV with:; as a separator;. representing missing values; no row
names included. Upload the .csv to TUWEL, together with the submission.
```{r}
write.csv2(df_vars, "df.csv", na = ".", row.names = FALSE)
```

